Equivalence Testing and the Second Generation \emph{P}-Value

### 2019-03-09 reply_to_reviews.R1.Rmd

3. Please do not feel like you ever need to apologize for having formal training we lack, that allows you to point out whenever we make formally incorrect statements. This is exactly why we submit our work for peer review - to be able to improve it based on feedback from experts. We now preface the sentence with "As long as data is normally distributed,". We follow your recommendation throughout to not generalize beyond the scenarios we study. In the second paragraph of the manuscript we now clearly state: "We limit our analysis to normally distributed continuous data." See also point 5 below.

4. After making the change at point 2 above, we now think the sentence "The conclusion of this section on asymmetric confidence intervals is that a SGPV of 1 or 0 can still be interpreted as a p < 0:025 or p > 0:975 in an equivalence test, since the SGPV and p-value for the TOST
procedure are always directly related at these values." should no longer be confusing. To make the sentence clearer, we have changed "these values" into "at the values p = 0.025 and p = 0.975" It should now be clear the sections are not contradictory - TOST differentiates below p < 0.025, when the SGPV is 1 for all p < 0.025.

5. We agree that is unclear whether the original statement by Blume et al that "the degree of overlap conveys how compatible the data are with the null premise" holds in general. We have now amended the sentence to read: "Although a SGPV of 1 or 0 can be directly interpreted, a SGPV between 0 and 1 is not interpretable as 'compatibility with the null hypothesis' under the assumption of normally distributed data, and the generalizability of this statement needs to be examined beyond normally distributed data requires further examination."

6. We have removed our earlier statements that the SGPV only works as a descriptive statistic. We deleted the sentence "the SGPV is a descriptive statistic (unlike the *p*-value that is calculated for an equivalence test, which is an inferential statistic) that provides the proportion of overlap of the confidence interval and the equivalence range." and deleted "The novelty of the SGPV lies in its use as a descriptive statistic" from the conclusion. Indeed, we acknowledge that "Although a SGPV of 1 or 0 can be directly interpreted, a SGPV between 0 and 1 is not interpretable as “compatibility with the null hypothesis”." So the SGPV is an inferential statistic when interpreted dichotomously. We will leave our uncertainty of how specific SGPV values (e.g., SGPV = 0.3) should be interpreted to future researchers, and we do not make any statements about whether such values are descriptive or inferential.

@@

2. The reviewer repeats that the correspondence only holds when SE's are small compared to the width of the equivalence range. The correspondence is not in the shape of the curve (which clearly differs across Figures 2, 6, and 9). We clearly state that: "When confidence intervals are symmetric we can think of the SGPV as a straight line that is directly related to the p-value from an equivalence test for three values." This correspondence is not dependent upon the SE's. It will always hold for continuous normal data. The reviewer says that "The paper does not provide theoretical justification for this connection nor any proofs." We are unsure which proof is needed for the overlap between the three points, since it is rather trivial and we believe the explanation in the paper is very clear.

For the relation between TOST p = 0.5 and SGPV = 0.5, TOST p = 0.025 and SGPV = 1, and TOST p = 0.975 and SGPV = 1: "The SGPV is 50% when the observed mean falls exactly on the lower or upper equivalence bound, because 50% of the symmetrical confidence interval overlaps with the equivalence range. When the observed mean equals the equivalence bound, the difference between the mean in the data and the equivalence bound is 0, the t-value for the equivalence test is also 0, and thus the p-value is 0.5 (situation A, Figure 3). Two other points always have to overlap. When the 95% CI falls completely inside the equivalence region, and one endpoint of the confidence interval is exactly equal to one of the equivalence bounds (see situation B in Figure 3) the TOST p-value (which relies on a one-sided test) is always 0.025, and the SGPV is 1. The third point where the SGPV and the p-value from the TOST procedure should overlap is where the 95% CI falls completely outside of the equivalence range, but one endpoint of the confidence interval is equal to the equivalence bound (see situation C in Figure 3), when the p-value will always be 0.975, and the SGPV is 0.

We do not think we can provide any other proof for this. A confidence interval and a p-value are directly related. The TOST procedure is a test using the confidence interval, and the SGPV (now that we clearly limited our discussion to a SGPV based on a confidence interval calculated from continuous normal data) compared against the same equivalence bounds will be related as we describe. But we believe our article might not have been clear enough about that we limit our discussion about the correspondence of the curves to these 3 points. Indeed, in some figures, the curves are less similar (e.g., Figure 6, 9, 13). Nevertheless, in these graphs, we still have the correspondence between these 3 points (or at least the 2 points if the CI is too wide compared to the equivalence bounds to ever observe a *p* = 0.025). To make it clearer that we talk about the overlap in these 3 critical points, we have added:

@@

3. We are aware of the Beger and Hsu (1996) article. But if the reviewer reads more closely, it should be clear that we never say the TOST is UMP. We say that TOST is a test procedure that aims to control error rates - our use of the term 'Neyman-Pearson perspective' refers to the classic difference between significance tests (Fisher) and hypothesis tests (Neyman-Pearson). We do not imply that any test that fixes an error rate is a NP test. Because many psychologists are not trained in the difference between a Fisherian approach and a Neyman Pearson approach to the use of p-values, we are simply adding a 'Neyman-Pearson' approach in the text to explain that "From a Neyman-Pearson perspective on statistical inferences the main conclusion is based on whether or not $p < \alpha$, and thus an equivalence test and SGPV can be performed by simply checking whether the confidence interval falls within the equivalence range, just as a null-hypothesis test can be performed by checking whether the confidence interval contains zero or not." We believe this is a correct use of the term 'Neyman-Pearson perspective on statistical inferences'.

4. We tried to find the passage the reviewer refers to when he writes "In light of this, the assertion without proof that classical p-values provide more or better assessments of the “evidence” appear out of place." but we could not. We do not use the word 'evidence' in the text (because neither the p-value from TOST nor the SGPV are measures of 'evidence' - we fully agree with BLume et al., 2018, in this respect). We also do not say the p-value is 'naturally informative'. Indeed, related to point 3 above, we tried to be consistent in interpreting the TOST procedure from a Neyman-Pearson perspective, and focus on error control. So we regrettably do not know how to incorporate this point.

5. We fully agree with the reviewer on point 5 that "The maximum Type I Error rate for the SGPV over the null
interval is 5%, exactly like the TOST test." Indeed, given the direct overlap between the TOST *p*-value and the SGPV at *p* = 0.025 and 0.0975 and SGPV of 1 and 0, any error rates (when concluding support for the null or alternative when the SGPV is either 1 or 0) must be identical. This is why we were so surprised when Blume et al (2018) write "They control the Type I error naturally, forcing it to zero as the sample size grows. This, in turn, offsets Type I Error inflation that results from multiple comparisons or multiple examinations of accumulating data. Findings identified by second-generation p-values are less likely to be false discoveries than findings identified by classical p-values. Consequently, second-generation p-values do not require ad-hoc adjustments to provide strict error control and this improves power in studies with massive multiple comparisons." Clearly, it can not both be true that the error rate of SGPV and TOST is identical, TOST requires adjusting for multiple comparisons, but the SGPV does not.

@@

Nevertheless, we agree this discussion was a bit too extensive. Given my own fondness of error control in statistical inferences, I get worried if researchers suggest we can ignore it, especially when making multiple comparisons. Imagine people start to use the SGPV instead of the TOST procedure whenever they perform 20 tests, and then argue they do not need to correct for multiple comparison - I think we can all agree this is undesireable! We have reduced the two paragraphs to one hopefully clear statement about the need to control error rates, also when using the SGPV. We hope this addresses the concerns by the editor as well as the reviewer, while still making an important point in the literature about the need to control error rates (if one cares about drawing incorrect conclusions).

@blume_second-generation_2018 claim that "Adjustments for multiple comparisons are obviated" (p. 15). However, this is not correct. Given the direct relationship between TOST and SGPV highlighted in this manuscript (where to TOST *p* = 0.025 equals SGPV = 1, as long as the SGPV is calculated based on confidence intervals, and assuming continuous and normally distributed data), not correcting for multiple comparisons will inflate the probability of concluding the absence of a meaningful effect based on the SGPV in exactly the same way as for equivalence tests. Whenever statistical tests are interpreted as support for a hypothesis (e.g., SPGV = 0 or SGPV =1), it is possible do so erroneously, and if researchers want to control error rates, they need to correct for multiple comparisons.

6. We have removed this section while addressing point 5 above, and thus incorporated the reviewers'suggestion to remove these statements.

7. We have not removed the paragraph about multiple comparisons, as the reviewer asks, but we have simplified our treatment of it. We do not need any other argument to make our claims about multiple comparisons that fact, noted by the reviewer, that "The maximum Type I Error rate for the SGPV over the null interval is 5%, exactly like the TOST test." This is true, and it logically follows (hence, no mathematical proof is needed) that error rates need to be controlled when using SGPV to conclude support for the null-hypothesis. We believe error control is important enough to include this paragraph in the paper, as we believe that lack of error control was one of the  causes of the replication crisis in psychology.

### 2019-03-09 manuscript.R1.Rmd
R1

abstract: |
  To move beyond the limitations of null-hypothesis tests, statistical approaches have been developed where the observed data are compared against a range of values that are equivalent to the absence of a meaningful effect. Specifying a range of values around zero allows researchers to statistically reject the presence of effects large enough to matter, and prevents practically insignificant effects from being interpreted as a statistically significant difference. We compare the behavior of the recently proposed second generation *p*-value [@blume_second-generation_2018] with the more established Two One-Sided Tests (TOST) equivalence testing procedure [@schuirmann_comparison_1987]. We show that the two approaches yield almost identical results under optimal conditions. Under suboptimal conditions (e.g., when the confidence interval is wider than the equivalence range, or when confidence intervals are asymmetric) the second generation *p*-value becomes difficult to interpret as a descriptive statistic. The second generation *p*-value is interpretable in a dichotomous manner (i.e., when the SGPV equals 0 or 1 because the confidence intervals lies completely within or outside of the equivalence range), but this dichotomous interpretation does not require calculations. We conclude that equivalence tests yield more consistent *p*-values, distinguish between datasets that yield the same second generation *p*-value, and allow for easier control of Type I and Type II error rates.

keywords          : "equivalence testing, second generation *p*-values, hypothesis testing, TOST, statistical inference" 
wordcount         : ""

@@

Second generation *p*-values (SGPV) were recently proposed to as a descriptive statistic that represents ‘the proportion of data-supported hypotheses that are also null hypotheses’ [@blume_second-generation_2018]. 
The researcher specifies an equivalence range around a null hypothesis of values that are considered practically equivalent to the null hypothesis. 
The SGPV measures the degree to which a set of data-supported parameter values falls within the interval null hypothesis. 
If the estimation interval falls completely within the equivalence range, the SGPV is 1. 
If the confidence interval falls completely outside of the equivalence range, the SGPV is 0. Otherwise the SGPV is a value between 0 and 1 that expresses the overlap of data-supported hypotheses and the equivalence range. 
When calculating the SGPV the set of data-supported parameter values can be represented by a confidence interval (CI), although one could also choose to use credible intervals or Likelihood support intervals (SI). 
When a confidence interval is used, the SGPV and equivalence tests such as the Two One-Sided Tests (TOST) procedure [@lakens_equivalence_2017;@meyners_equivalence_2012;@quertemont_how_2011;@schuirmann_comparison_1987] appear to have close ties, because both tests compare a confidence interval against an equivalence range.
Here, we aim to examine the similarities and differences between the TOST procedure and the SGPV. 
We limit our analysis to normally distributed continuous data.

@@

We can examine this situation by calculating the SGPV and performing the TOST for a situation where sample sizes are small and the equivalence range is narrow, such that the CI is more than twice as large as the equivalence range (see Figure \ref{fig:TOSTSGPV6}). When the two statistics are plotted against each other we can see where they are unrelated (indicated by straight lines in the curve, see Figure \ref{fig:TOSTSGPV7}). We see the SGPV is 0.5 for a range of observed means where the *p*-value from the equivalence test still varies. It should be noted that in these calculations the *p*-values for the TOST procedure are *never* smaller than 0.05 (i.e., they do not get below 0.05 on the y-axis). In other words, we cannot conclude equivalence based on any of the observed means. This happens because the TOST *p*-value is smaller than 0.05 only when the 90% CI falls completely between the upper and lower equivalence bounds. However, we are examining a scenario where the 90% CI is so wide that it never falls completely within the two equivalence bounds, and thus the equivalence test is never significant. As @lakens_equivalence_2017 notes: “in small samples (where CIs are wide), a study might have no statistical power (i.e., the CI will always be so wide that it is necessarily wider than the equivalence bounds).” None of the *p*-values based on the TOST procedure are below 0.05, and thus, in the long run we have 0% power.

@@

The *p*-value from the TOST procedure and the SGPV are also unrelated when the CI is wider than the equivalence range (so the precision is low) and overlaps with the upper and lower equivalence bound, but the CI is *not* twice as wide as the equivalence range. In the example below, we see that the CI is only 1.79 times as wide as the equivalence bounds, but the CI overlaps with the lower and upper equivalence bounds (Figure \ref{fig:TOSTSGPV8}). This means the SGPV is not set to 0.5, but it is constant across a range of observed means.

@@

If we once more plot the two statistics against each other to see where they are unrelated (indicated by straight lines in the curve), we see the SGPV is 0.56 for a range of observed means where the *p*-value from the equivalence test still varies (Figure \ref{fig:TOSTSGPV10}).

@@

To conclude this section, there are three situations where the *p*-value from the TOST procedure is unrelated to the SGPV. In all these situations the *p*-value for the equivalence test differentiates tests with different means, but the SGPV does not. Therefore, as a purely continuous descriptive statistic, the SGPV is more limited than the *p*-value from the TOST procedure.

@@

This example illustrates the difference between a proportion and a probability. As long as data are normally distributed, there is always a 50% probability of observing a correlation smaller or larger than the true correlation, but the SGPV for this situation depends on how far away the observed correlation is from 0. The further away from 0, the larger the SGPV when the observed mean falls on the equivalence bound. The SGPV is the proportion of values in a 95% confidence interval that overlap with the equivalence range, but not the probability that these values will be observed. In the most extreme case (i.e., a sample size of 4, and equivalence bounds set to r = -0.99 and 0.99, with an observed correlation of 0.99) `r round(sgpv,3)*100`% of the confidence interval overlaps with the equivalence range, even though in the long run only 50% of the correlations observed in the future will fall in this range. It should be noted that in larger sample sizes the SGPV is closer to 0.5 whenever the observed correlation falls on the equivalence bound, but this extreme example nevertheless clearly illustrates the difference between question the SGPV answers, and the question a *p*-value answers. The conclusion of this section on asymmetric confidence intervals is that a SGPV of 1 or 0 can still be interpreted as a *p* < 0.025 or *p* > 0.975 in an equivalence test, since the SGPV and *p*-value for the TOST procedure are always directly related at the values *p* = 0.025 and *p* = 0.975. Although @blume_second-generation_2018 state that "the degree of overlap conveys how compatible the data are with the null premise" this definition of what the SGPV provides does not hold for asymmetric confidence intervals. Although a SGPV of 1 or 0 can be directly interpreted, a SGPV between 0 and 1 is not interpretable as 'compatibility with the null hypothesis' under the assumption of normally distributed data, and the generalizability of this statement needs to be examined beyond normally distributed data. Indeed, Blume and colleagues write in the supplemental material that "The magnitude of an inconclusive second-generation *p*-value can vary slightly when the effect size scale is transformed. However definitive findings, i.e. a *p*-value of 0 or 1 are *not* affected by the scale changes."

@@

@blume_second-generation_2018 claim that "Adjustments for multiple comparisons are obviated" (p. 15). However, this is not correct. Given the direct relationship between TOST and SGPV highlighted in this manuscript (where to TOST *p* = 0.025 equals SGPV = 1, as long as the SGPV is calculated based on confidence intervals, and assuming continuous and normally distributed data), not correcting for multiple comparisons will inflate the probability of concluding the absence of a meaningful effect based on the SGPV in exactly the same way as for equivalence tests. Whenever statistical tests are interpreted as support for a hypothesis (e.g., SPGV = 0 or SGPV =1), it is possible do so erroneously, and if researchers want to control error rates, they need to correct for multiple comparisons.

>>>>

abstract: |
  To move beyond the limitations of null-hypothesis tests, statistical approaches have been developed where the observed data are compared against a range of values that are equivalent to the absence of a meaningful effect. Specifying a range of values around zero allows researchers to statistically reject the presence of effects large enough to matter, and prevents practically insignificant effects from being interpreted as a statistically significant difference. We compare the behavior of the recently proposed second generation *p*-value [@blume_second-generation_2018] with the more established Two One-Sided Tests (TOST) equivalence testing procedure [@schuirmann_comparison_1987]. We show that the two approaches yield almost identical results under optimal conditions. Under suboptimal conditions (e.g., when the confidence interval is wider than the equivalence range, or when confidence intervals are asymmetric) the second generation *p*-value becomes difficult to interpret. The second generation *p*-value is interpretable in a dichotomous manner (i.e., when the SGPV equals 0 or 1 because the confidence intervals lies completely within or outside of the equivalence range), but this dichotomous interpretation does not require calculations. We conclude that equivalence tests yield more consistent *p*-values, distinguish between datasets that yield the same second generation *p*-value, and allow for easier control of Type I and Type II error rates.

keywords          : "equivalence testing, second generation p-values, hypothesis testing, TOST, statistical inference"

@@

Second generation *p*-values (SGPV) were recently proposed to as a statistic that represents ‘the proportion of data-supported hypotheses that are also null hypotheses’ [@blume_second-generation_2018]. 
The researcher specifies an equivalence range around a null hypothesis of values that are considered practically equivalent to the null hypothesis. 
The SGPV measures the degree to which a set of data-supported parameter values falls within the interval null hypothesis. 
If the estimation interval falls completely within the equivalence range, the SGPV is 1. 
If the confidence interval falls completely outside of the equivalence range, the SGPV is 0. Otherwise the SGPV is a value between 0 and 1 that expresses the overlap of data-supported hypotheses and the equivalence range. 
When calculating the SGPV the set of data-supported parameter values can be represented by a confidence interval (CI), although one could also choose to use credible intervals or Likelihood support intervals (SI). 
When a confidence interval is used, the SGPV and equivalence tests such as the Two One-Sided Tests (TOST) procedure [@lakens_equivalence_2017;@meyners_equivalence_2012;@quertemont_how_2011;@schuirmann_comparison_1987] appear to have close ties, because both tests compare a confidence interval against an equivalence range.
Here, we aim to examine the similarities and differences between the TOST procedure and the SGPV. 
We limit our analysis to continuous data sampled from a bivariate normal distribution.

@@

We can examine this situation by calculating the SGPV and performing the TOST for a situation where sample sizes are small and the equivalence range is narrow, such that the CI is more than twice as large as the equivalence range (see Figure \ref{fig:TOSTSGPV6}). When the two statistics are plotted against each other we can see where the SGPV is the same while the TOST *p*-value still differentiates different observed means (indicated by straight lines in the curve, see Figure \ref{fig:TOSTSGPV7}). We see the SGPV is 0.5 for a range of observed means where the *p*-value from the equivalence test still varies. It should be noted that in these calculations the *p*-values for the TOST procedure are *never* smaller than 0.05 (i.e., they do not get below 0.05 on the y-axis). In other words, we cannot conclude equivalence based on any of the observed means. This happens because the TOST *p*-value is smaller than 0.05 only when the 90% CI falls completely between the upper and lower equivalence bounds. However, we are examining a scenario where the 90% CI is so wide that it never falls completely within the two equivalence bounds, and thus the equivalence test is never significant. As @lakens_equivalence_2017 notes: “in small samples (where CIs are wide), a study might have no statistical power (i.e., the CI will always be so wide that it is necessarily wider than the equivalence bounds).” None of the *p*-values based on the TOST procedure are below 0.05, and thus, in the long run we have 0% power.

@@

The *p*-value from the TOST procedure still differentiates observed means, while the  SGPV does not, when the CI is wider than the equivalence range (so the precision is low) and overlaps with the upper and lower equivalence bound, but the CI is *not* twice as wide as the equivalence range. In the example below, we see that the CI is only 1.79 times as wide as the equivalence bounds, but the CI overlaps with the lower and upper equivalence bounds (Figure \ref{fig:TOSTSGPV8}). This means the SGPV is not set to 0.5, but it is constant across a range of observed means, while the TOST *p*-value is not constant across this range.

@@

If we once more plot the two statistics against each other we see the SGPV is 0.56 for a range of observed means where the *p*-value from the equivalence test still varies, as indicated by the straight section of the line (Figure \ref{fig:TOSTSGPV10}).

@@

To conclude this section, there are situations where the *p*-value from the TOST procedure continues to differentiate, while the SGPV does not. Therefore, interpreted as a continuous statistic, the SGPV is more limited than the *p*-value from the TOST procedure.

@@

This example illustrates the difference between a proportion and a probability. As long as data are sampled from a bivariate normal distribution, there is always a 50% probability of observing a correlation smaller or larger than the true correlation, but the SGPV for this situation depends on how far away the observed correlation is from 0. The further away from 0, the larger the SGPV when the observed mean falls on the equivalence bound. The SGPV is the proportion of values in a 95% confidence interval that overlap with the equivalence range, but not the probability that these values will be observed. In the most extreme case (i.e., a sample size of 4, and equivalence bounds set to r = -0.99 and 0.99, with an observed correlation of 0.99) `r round(sgpv,3)*100`% of the confidence interval overlaps with the equivalence range, even though in the long run only 50% of the correlations observed in the future will fall in this range. It should be noted that in larger sample sizes the SGPV is closer to 0.5 whenever the observed correlation falls on the equivalence bound, but this extreme example nevertheless clearly illustrates the difference between question the SGPV answers, and the question a *p*-value answers. The conclusion of this section on asymmetric confidence intervals is that a SGPV of 1 or 0 can still be interpreted as a *p* < 0.025 or *p* > 0.975 in an equivalence test, since the SGPV and *p*-value for the TOST procedure are always directly related at the values *p* = 0.025 and *p* = 0.975. Although @blume_second-generation_2018 state that "the degree of overlap conveys how compatible the data are with the null premise" this definition of what the SGPV provides does not hold for asymmetric confidence intervals. Although a SGPV of 1 or 0 can be directly interpreted, a SGPV between 0 and 1 is not interpretable as 'compatibility with the null hypothesis' under the assumption of a bivariate normal distribution, and the generalizability of this statement needs to be examined beyond normal bivariate distributions. Indeed, Blume and colleagues write in the supplemental material that "The magnitude of an inconclusive second-generation *p*-value can vary slightly when the effect size scale is transformed. However definitive findings, i.e. a *p*-value of 0 or 1 are *not* affected by the scale changes."

@@

@blume_second-generation_2018 claim that "Adjustments for multiple comparisons are obviated" (p. 15). However, this is not correct. Given the direct relationship between TOST and SGPV highlighted in this manuscript (where to TOST *p* = 0.025 equals SGPV = 1, as long as the SGPV is calculated based on confidence intervals, and assuming data sampled from a continuous bivariate normal distribution), not correcting for multiple comparisons will inflate the probability of concluding the absence of a meaningful effect based on the SGPV in exactly the same way as for equivalence tests. Whenever statistical tests are interpreted as support for a hypothesis (e.g., SPGV = 0 or SGPV =1), it is possible do so erroneously, and if researchers want to control error rates, they need to correct for multiple comparisons.
